A Parallel General Purpose Multi-Objective Optimization Framework, with Application To Beam Dynamics

### 2018-07-12 paper/prab.tex
starting to fix verbatim sections
fixing figures of tba beam line
edits to timing and hyper parameter sections
first round of figures

To allow arbitrary constraints and objective expressions, such as
%
\vspace{0.2cm}
\begin{Verbatim}[fontsize=\scriptsize]
  name: OBJECTIVE,
        EXPR="5 * average(42.0, "measurement.dat") + ENERGY";
\end{Verbatim}
\vspace{0.2cm}
%
\noindent
  we implemented an expression parser using Boost
%  Spirit\footnote{\url{http://boost-spirit.com/}}.
In addition to the parser, we need an evaluator able to evaluate an expression,
  given a parse tree and variable assignments to an actual value.
Expressions arising in multi-objective optimization problems usually evaluate
  to booleans or floating point values.
The parse tree, also denoted abstract syntax tree (AST), is constructed
  recursively while an expression is parsed.
Upon evaluation, all unknown variables are replaced with values, either
  obtained from simulation results or provided by other subtrees in the AST.
In this stage, the AST can be evaluated bottom-up and the desired result is
  returned after processing the root of the tree.

@@

If the objectives and constraints are simple arithmetical expressions, the
  \texttt{FunctionEvaluator} simulator can be used.
Using functors and the default expression primitives already powerful
  multi-objective optimization problems can be specified, i.e.\ the benchmark
  problem presented in \cite{hbwh:05}:
%
\small{
\begin{align} \label{eqn:bench}
  \text{ min} & \left[ 1 - \exp \left( -1 \left(
    \left(x_1 - \frac{1}{\sqrt{3}} \right)^2 +
    \left(x_2 - \frac{1}{\sqrt{3}} \right)^2 +
    \left(x_3 - \frac{1}{\sqrt{3}} \right)^2 \right)\right), \right. \\
                    & \quad \left. 1 - \exp \left( -1 \left(
    \left(x_1 + \frac{1}{\sqrt{3}} \right)^2 +
    \left(x_2 + \frac{1}{\sqrt{3}} \right)^2 +
    \left(x_3 + \frac{1}{\sqrt{3}} \right)^2 \right)\right) \right]^T \nonumber \\
    \text{s.t.} \quad & \quad -1 \le x_i \le 1, \quad i=1,2,3 \nonumber
  \text{.}
\end{align}
}
%
Using the default expression primitives this can be stated in an input file
  as:

\vspace{0.2cm}
\begin{flushleft}
\noindent
\begin{Verbatim}[fontsize=\scriptsize]
d1: DVAR, VARIABLE="x1", LOWERBOUND="-1.0", UPPERBOUND="1.0";
d2: DVAR, VARIABLE="x2", LOWERBOUND="-1.0", UPPERBOUND="1.0";
d3: DVAR, VARIABLE="x3", LOWERBOUND="-1.0", UPPERBOUND="1.0";

@@

objs:    OBJECTIVES = (obj1, obj2);
dvars:   DVARS = (d1, d2, d3);
constrs: CONSTRAINTS = ();
opt: OPTIMIZE, OBJECTIVES=objs, DVARS=dvars, CONSTRAINTS=constrs;
\end{Verbatim}
\end{flushleft}

\vspace{0.2cm}

@@

Interestingly, the computation of a good approximation of the Pareto front is
  already feasible on a small cluster using only a modest number of MPI
  procsses.
For 1000 generations consisting of 2048 feasible and 8354 infeasible function
  evaluations (simulations) the optimizer needed $845$ minutes on the
  \textsc{Felsim} cluster on $16$ processes.
When we double the number of processes to $32$ (\textsc{Felsim} allows a
  maximum of $64$ processes per job) the runtime improves to $302$ minutes (2048
  accepted and 5426 infeasible individuals).
Notice that the runtime can vary due to the random nature of the process.
Nevertheless we see a reasonable speeduo when using more cores to solve the
  optimization problem and the parallel performance will be further evaluated
  in a forthcoming publication.
%The number of individuals is the only limiting factor.

@@

The authors thank the \textsc{SwissFEL} and AWA teams for contributing to the
  formulation of optimization problems. 
  We gratefully acknowledge the computing resources provided on Bebop,
  a HPC cluster operated by the LCRC at ANL.
  This work was partly supported by the 
  U.S. Department of Energy, Office of Science, under 
  contract number DE-AC02-06CH11357 and grant number DE-SC0015479.

>>>>

To allow arbitrary constraints and objective expressions, such as
%
%{0.2cm}
\begin{Verbatim}[fontsize=\scriptsize]
  name: OBJECTIVE,
        EXPR="5 * average(42.0, "measurement.dat") + ENERGY";
\end{Verbatim}
%\vspace{0.2cm}
%
\noindent
  we implemented an expression parser using Boost
%  Spirit\footnote{\url{http://boost-spirit.com/}}.
In addition to the parser, we need an evaluator able to evaluate an expression,
  given a parse tree and variable assignments to an actual value.
Expressions arising in multi-objective optimization problems usually evaluate
  to booleans or floating point values.
The parse tree, also denoted abstract syntax tree (AST), is constructed
  recursively while an expression is parsed.
Upon evaluation, all unknown variables are replaced with values, either
  obtained from simulation results or provided by other subtrees in the AST.
In this stage, the AST can be evaluated bottom-up and the desired result is
  returned after processing the root of the tree.

@@

If the objectives and constraints are simple arithmetical expressions, the
  \texttt{FunctionEvaluator} simulator can be used.
Using functors and the default expression primitives already powerful
  multi-objective optimization problems can be specified, i.e.\ the benchmark
  problem presented in \cite{hbwh:05}:
%

\begin{flushleft}
\begin{align} \label{eqn:bench}
\text{min} & \left[ 1 - \exp \left( -1 \left(
\left(x_1 - \frac{1}{\sqrt{3}} \right)^2 +
\left(x_2 - \frac{1}{\sqrt{3}} \right)^2 +
\left(x_3 - \frac{1}{\sqrt{3}} \right)^2 \right)\right), \right. \\
& \left. 1 - \exp \left( -1 \left(
\left(x_1 + \frac{1}{\sqrt{3}} \right)^2 +
\left(x_2 + \frac{1}{\sqrt{3}} \right)^2 +
\left(x_3 + \frac{1}{\sqrt{3}} \right)^2 \right)\right) \right]^T \nonumber \\
\text{s.t.} \quad & \quad -1 \le x_i \le 1, \quad i=1,2,3 \nonumber
\text{.}
\end{align}
\end{flushleft}

%
Using the default expression primitives this can be stated in an input file
  as:
%
%\vspace{0.2cm}

\begin{flushleft}
\begin{Verbatim}[fontsize=\scriptsize]
d1: DVAR, VARIABLE="x1", LOWERBOUND="-1.0", UPPERBOUND="1.0";
d2: DVAR, VARIABLE="x2", LOWERBOUND="-1.0", UPPERBOUND="1.0";
d3: DVAR, VARIABLE="x3", LOWERBOUND="-1.0", UPPERBOUND="1.0";

@@

objs:    OBJECTIVES = (obj1, obj2);
dvars:   DVARS = (d1, d2, d3);
constrs: CONSTRAINTS = ();
opt: OPTIMIZE, OBJECTIVES=objs, DVARS=dvars, CONSTRAINTS=constrs;
\end{Verbatim}
\end{flushleft}
%\vspace{0.2cm}

@@

Interestingly, the computation of a good approximation of the Pareto front is
  already feasible on a small cluster using only a modest number of MPI
  processes.
For 1000 generations consisting of 2048 feasible and 8354 infeasible function
  evaluations (simulations) the optimizer needed $845$ minutes on the
  \textsc{Felsim} cluster on $16$ processes.
When we double the number of processes to $32$ (\textsc{Felsim} allows a
  maximum of $64$ processes per job) the runtime improves to $302$ minutes (2048
  accepted and 5426 infeasible individuals).
Notice that the runtime can vary due to the random nature of the process.
Nevertheless we see a reasonable speeduo when using more cores to solve the
  optimization problem. %and the parallel performance will be further evaluated
%  in a forthcoming publication.
%The number of individuals is the only limiting factor.

The framework was then applied to a larger and untested beam dynamics problem at the AWA.
Optimization of the 3D beam size and energy spread was accomplished.
A scan of time step and hyper parameters was done to determine computational settings.
Then a full scale physics optimization was performed.
The resulting beam tune can and will be used in future TBA experiments at the AWA.
This problem also took advantage of the frameworks massively parallel capabilities by
running on 2624 processes on Bebop at LCRC.

@@

The authors thank the \textsc{SwissFEL} and AWA teams for contributing to the
  formulation of optimization problems. 
  We gratefully acknowledge the computing resources provided on Bebop,
  a HPC cluster operated by the LCRC at ANL.
  Thanks to Scott Doran for providing CAD drawings of the AWA beam lines.
  This work was partly supported by the 
  U.S. Department of Energy, Office of Science, under 
  contract number DE-AC02-06CH11357 and grant number DE-SC0015479.

### 2018-07-12 paper/theor-experiment.tex
first round of figures

After collection of the data for all four experiments, several metrics
were compared, including number of generations completed in twenty four hours and
Pareto fronts at $s_1$ and $s_2$.
From Table \ref{extable}, we clearly see ex-3 is significantly 
slower, as it evaluated only 53 generations compared to the max 
experiment, ex-1 at 96 generations.
Perhaps this trade off would be acceptable if the Pareto front was significantly 
improved, but from Fig. \ref{expareto}, we can see this is not the case.
Similar arguments can be made for ex-2, which evaluated about 15 less generations
with no clear benefit in either Pareto front.
The Pareto fronts at $s_1$, are nearly identical. We could expect
this trend would continue given more time. 
When looking at the Pareto front at $s_2$, only ex-4 has a slight 
edge over the others.
With ex-2 and ex-3 eliminated due to evaluation time, 
and a slightly better Pareto front at $s_2$ for ex-4, the hyper parameters in ex-4 were chosen as the default values for subsequent runs.
\iffalse  
\begin{figure}
	\centering
	\includegraphics[width=0.75\textwidth]{Report/ex-pareto1}%
	\llap{\raisebox{1.8cm}{%  move next graphics to top right corner
			\includegraphics[height=5.5cm]{Report/ex-pareto2}%
	}}
	\caption{My overlay figure.}
\end{figure}  
\fi

\begin{figure}
	\centering
	\begin{minipage}{0.49\textwidth}
		\centering
		\includegraphics[width=\textwidth]{Report/ex-pareto1}
		(a) Pareto fronts for ex-1 through ex-4 at $s_1$.
		%\caption{(a) Pareto fronts at $s_1$.}
		%\label{expareto1}
	\end{minipage}
	\begin{minipage}{0.49\textwidth}
		\centering
		\includegraphics[width=\textwidth]{Report/ex-pareto2}
		(b) Pareto fronts for ex-1 through ex-4 at $s_2$.
		%\caption{(b) Pareto fronts at $s_2$.}
		%\label{expareto2}
	\end{minipage}
	\caption{Comparison of Pareto fronts for initial optimization experiments, ex-1 through ex-4.}
	\label{expareto}
\end{figure}

@@

\begin{figure}
	\includegraphics[width=0.5\textwidth]{../pareto_stat_plots/y_vs_py_pareto_front_quads_before_Q5}
	\caption{Pareto front comparing longitudinal beam size ($\sigma_z$) and energy spread ($dE$).}
	\label{fig:pareto2}
\end{figure}

As expected, the x dimension is impacted by the bending elements, and unable to maintain 
the small beam sizes seen in the y dimension. 
The $\sigma_x$ and $\sigma_{px}$ plots are of most interest, as these parameters are the heavily 
impacted by bending elements. With this in mind, several beam parameters corresponding to
options on Pareto Front in Fig. X were plotted and compared. 
The best results are shown in Fig. Z. 
The maximum beam sizes are well below the beam pip aperture limits as shown in Fig. (b).

>>>>

After collection of the data for all four experiments, several metrics
were compared, including number of generations completed in twenty four hours and
Pareto fronts at $s_1$ and $s_2$.
From Table \ref{extable}, we clearly see ex-3 is significantly 
slower, as it evaluated only 53 generations compared to the max 
experiment, ex-1 at 96 generations.
Perhaps this trade off would be acceptable if the Pareto front was significantly 
improved, but from Fig. \ref{expareto}, we can see this is not the case.
Similar arguments can be made for ex-2, which evaluated about 15 less generations
with no clear benefit in either Pareto front.
The Pareto fronts at $s_1$, are nearly identical. We could expect
this trend would continue given more time. 
When looking at the Pareto front at $s_2$, only ex-4 has a slight 
edge over the others.
With ex-2 and ex-3 eliminated due to evaluation time, 
and a slightly better Pareto front at $s_2$ for ex-4, the hyper parameters in ex-4 were chosen as the default values for subsequent runs.

\begin{figure}
	\centering
	\begin{minipage}{0.49\textwidth}
		\centering
		\includegraphics[width=\textwidth]{Report/ex-pareto1}
		(a) Pareto fronts for ex-1 through ex-4 at $s_1$.
		%\caption{(a) Pareto fronts at $s_1$.}
		%\label{expareto1}
		\vspace{2em}
	\end{minipage}
	\begin{minipage}{0.49\textwidth}
		\centering
		\includegraphics[width=\textwidth]{Report/ex-pareto2}
		(b) Pareto fronts for ex-1 through ex-4 at $s_2$.
		%\caption{(b) Pareto fronts at $s_2$.}
		%\label{expareto2}
	\end{minipage}
	\caption{Comparison of Pareto fronts for initial optimization experiments, ex-1 through ex-4.}
	\label{expareto}
\end{figure}

@@

\begin{figure}
	\centering
	\includegraphics[width=0.48\textwidth]{../pareto_stat_plots/dE_vs_zrms_pareto_front_quads_before_Q5_zoomout}%
	\llap{\raisebox{1.9cm}{%  move next graphics to top right corner
			\includegraphics[height=4.5cm]{../pareto_stat_plots/dE_vs_zrms_pareto_front_quads_before_Q5}%
	}}
	\caption{Comparison of energy spread and bunch length ($\sigma_z$). \nrnote{add tikz arrows to point to left side of curve?}}
	\label{fig:pareto2}
\end{figure}

As expected, the x dimension is impacted by the bending elements, and unable to reach 
the small beam sizes seen in the y dimension. This suggest objectives in the x 
should be weighted higher than y when choosing operating parameters. 
In the case of bunch length, there are not many options to choose from.
Ignoring the extreme value on the right, Fig.~\ref{fig:pareto2} shows the 
range of optimal bunch lengths fall between 1.39 and 1.5 mm.

With these observations in mind, several beam parameters corresponding to
options on Pareto Front in Fig.~\ref{fig:pareto1} were plotted and compared. 
A select result is shown in Fig.~\ref{fig:stat}. 
The maximum beam sizes are well below the beam pipe aperture limits as shown in Fig.~\ref{fig:stat} (b).
The solution is nearly round, which will increase changes of keeping the beam nearly round
as it travels to the last triplet in Fig.~\ref{awa-tba}.
Overall this solution is satisfactory, and meets all requirements at the AWA.

\begin{figure}
	\includegraphics[width=0.48\textwidth]{../pareto_stat_plots/xyrms-zrms_newobj}\\
	\includegraphics[width=0.48\textwidth]{../pareto_stat_plots/xy-max-min-zrms_newobj}
	\caption{Optimized beam sizes along high charge beam line.}
	\label{fig:stat}
\end{figure}
